---
title: Accelerated component-wise gradient boosting using efficient data representation and momentum-based optimization
author: D. Schalk, B. Bischl, D. Rügamer
date: 17th, Dec. 2022
output:
  beamer_presentation:
    keep_tex: true
    includes:
      in_header: "style/preamble_reisensburg.sty"
    template: "style/custom_pandoc.tex"
---

```{r setup, include=FALSE}
options(width = 80)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/"
)

library(ggplot2)
library(ggsci)
library(ggthemes)
library(patchwork)

set.seed(31415)

matToTex = function(mat) {
  rows = paste0("\t", apply(mat, MARGIN = 1, paste, collapse = " & "))
  matrix_string = paste(rows, collapse = " \\\\\n ")
  return(paste("\\begin{pmatrix}\n", matrix_string, "\n\\end{pmatrix}"))
}

ggplot2::theme_set(theme_few())
scale_colour_discrete = function(...) scale_color_aaas(...)

x = c(15, 20, 25, 30, 50, 70, 84)
y = c(8,   0, -4,  0,  4,  6,  5)

m1 = lm(y ~ poly(x, 4))

## SIMULATION:
nsim = 1000L
xage = sample(seq(15, 84), nsim, replace = TRUE)
fage = predict(m1, data.frame(x = xage))
xcountry = sample(c("DE", "AT", "GB", "ES", "FR"), nsim, replace = TRUE)
ecountry = c(DE = -2, AT = 4, GB = -1, ES = 2, FR = 0)
fcountry = ecountry[xcountry]
xincome = sample(seq(700, 5000), nsim, replace = TRUE)

yhappiness = 2 + fage + fcountry + rnorm(nsim, 0, 1)
r0 = yhappiness - mean(yhappiness)

library(cpsp)

knotsincome = createKnots(xincome, 10, 3)
Zincome = createSplineBasis(xincome, 3, knotsincome)
Kincome = penaltyMat(nparams = ncol(Zincome), differences = 2)
penincome = demmlerReinsch(t(Zincome) %*% Zincome, Kincome, 5)
bincome = solve(t(Zincome) %*% Zincome + penincome * Kincome) %*% t(Zincome) %*% r0

knotsage = createKnots(xage, 10, 3)
Zage = createSplineBasis(xage, 3, knotsage)
Kage = penaltyMat(nparams = ncol(Zage), differences = 2)
penage = demmlerReinsch(t(Zage) %*% Zage, Kage, 5)
bage = solve(t(Zage) %*% Zage + penage * Kage) %*% t(Zage) %*% r0

oage = order(xage)
oincome = order(xincome)

dfincome = data.frame(x = sort(xincome), y = Zincome[oincome, ] %*% bincome)
dfage = data.frame(x = sort(xage), y = Zage[oage, ] %*% bage)
dfcountry = data.frame(x = names(ecountry), y = unname(ecountry))

ggage = ggplot(dfage, aes(x, y)) + geom_line() + xlab("age") + ylab(expression(b[age](x)))
ggincome = ggplot(dfincome, aes(x, y)) + geom_line() + xlab("income") + ylab(expression(b[income](x)))
ggcountry = ggplot(dfcountry, aes(x, y)) + geom_boxplot() + xlab("country") + ylab(expression(b[country](x)))
```

# Background

## About the project

- Component-wise boosting \citep[CWB;][]{buhlmann2003boosting,buhlmann2007boosting} is a boosting method based on multiple base learners.
- Utilizing interpretable statistical models as base learners makes the fitted model interpretable.
- Also, CWB is can fit models in high dimensional features spaces ($n \ll p$) and provides an inherent (unbiased) feature selection.
- But, depending on the size of the data, it is infeasible to fit the algorithm w.r.t runtime or memory allocation.

\begin{itemize}
\item[$\Rightarrow$] \textbf{Goal:} Fasten the fitting process and reduce the memory consumption without loosing predictive or estimation accuracy.
\end{itemize}

## CWB basics I

- Feature vector $\bm{x} = (x_1, \dots, x_p)\in\mathcal{X}$
- CWB fits a model $\fh$ based on an additive structure $$\mathbb{E}[Y|\bm{x}] = \fh(\bm{x}) = f_0 + \sum_{k=1}^K b_k(\bm{x}).$$ Hence, CWB can be used as fitting engine for GAMs.
- $K$ additive terms represented by base learner $b_k : \mathcal{X} \to \mathbb{R}$. Often one base learner per feature is used to model univariate effects.


## CWB basics II

- Example throughout the talk:
    - Three features $x_{\text{age}}$, $x_{\text{country}}$, and $x_{\text{income}}$
    - Response $Y$ is a numeric score for "happiness"
    - Aim: Model $Y$ with $b_{\text{age}}$ a P-spline, $b_{\text{income}}$ a P-spline and $b_{\text{country}}$ a one-hot encoding

```{r, echo=FALSE, fig.width=9, fig.height=2, out.widht="0.9\textwidth", fig.align="center"}
ggage + ggincome + ggcountry
```

## CWB algorithm I

- Given is a data set $\D$ with $n$ observations and $p$ features, a loss function $L(y, \hat{y})$, and a set of base learners $b_1, \dots, b_\blK$.
- The gaol of fitting CWB is to minimize the empirical risk $\riske(\fh | \D) = n^{-1}\sum_{(x,y)\in\D} L(y, \fh(x))$.
- The algorithm is initialized with a loss-optimal constant $c$ that minimizes the empirical risk: $$f_0 = \fh^{[0]}(\xv) = \argmin_{c\in\mathcal{Y}}\riske(c|\D)$$
- Model updates are calculated by functional gradient descent: $$\fh^{[m+1]} = \fmh + \nu \hat{b}^{[m]}$$

## CWB algorithm II

- The negative functional gradient is expressed by pseudo residuals: $$\rmi = -\nabla_f L(y^{(i)}, f(\xi))|_{f = \fmdh},\ i \in \{1, \dots, n\}$$
- In CWB, each base learner $b_1, \dots, b_K$ is fitted to $\rmm$ and the one with the lowest sum of squared errors (SSE) is chosen as new component $\hat{b}^{[m]}$.
- The last two step are repeated $M$ times.


## CWB algorithm III

```{=latex}
\begin{algorithm}[H]

\footnotesize

\caption{Vanilla CWB algorithm}\label{algo:cwb}
\vspace{0.15cm}
\hspace*{\algorithmicindent} \textbf{Input} Train data $\D$, learning rate $\nu$, number of boosting iterations $M$, loss\\
\hspace*{\algorithmicindent} \phantom{\textbf{Input} } function $L$, base learners $b_1, \dots, b_\blK$\\
\hspace*{\algorithmicindent} \textbf{Output} Model $\fh = \hat{f}^{[M]}$, estimated coefficient vectors $\tbih{1}, \ldots, \tbih{M}$\vspace{0.15cm}
\hrule

\begin{algorithmic}[1]
\Procedure{$\operatorname{CWB}$}{$\D,\nu,L,b_1, \dots, b_\blK$}
    \State Initialize: $f_0 = \fh^{[0]}(\xv) = \argmin_{c\in\mathcal{Y}}\riske(c|\D)$
    \While{$m \leq M$}
        \State $\rmi = -\left.\pd{\Lxyi}{f(\xi)}\right|_{f = \fmdh},\ \ \forall i \in \{1, \dots, n\}$%\label{algo:cwb:line:pr}
        \For{$\blk \in \{1, \dots, \blK\}$}
            \State $\tbmh_\blk = \left(\design_\blk^\tran \design_\blk + \bm{K}_\blk\right)^{-1} \design^\tran_\blk \rmm$%\label{algo:cwb:line:fitbl}
            \State $\sse_\blk = \sum_{i=1}^n(\rmi - b_\blk(\xi | \tbmh_\blk))^2$% \label{algo:cwb:line:sse}
        \EndFor
        \State $\blk^{[m]} = \argmin_{\blk\in\{1, \dots, \blK\}} \sse_\blk$% \label{algo:cwb:line:blselection}
        \State $\fmh(\xv) = \fmdh(\xv) + \nu b_{\blk^{[m]}} (\xv | \tbmh_{\blk^{[m]}})$
    \EndWhile
    \State \textbf{return} $\fh = \fh^{[M]}$
\EndProcedure
\end{algorithmic}
\end{algorithm}
```

## Goal of the project

$\Rightarrow$ Increase CWB's efficiency:

- __Acceleration:__ Fasten the fitting process by using Nesterovs momentum.
- __Memory:__ Reduce the memory consumption by discretizing numerical features.

# Acceleration

## Nesterovs momentum

__Gradient descent:__

```{=latex}
\vspace{0.2cm}
{\small
\begin{tabular}{ccc}
  Parameter space & & Functional space \\[0.3cm]
  $\tbh^{[m+1]} = \tbh^{[m]} + \nu \nabla_{\tb}\riske(\fh(. | \tbh^{[m]}) | \D)$ & $\Rightarrow$ & $\fh^{[m+1]} = \fmh + \nu \hat{b}^{[m]}$
\end{tabular}}
\vspace{0.2cm}
```

__Nesterov momentum:__


```{=latex}
\vspace{0.2cm}
{\small
\begin{tabular}{ccc}
  Parameter space & & Functional space \\[0.3cm]
  $\bm{u}^{[m]} = \nabla_{\tb}\riske(\fh(. | \tbh^{[m]} - \gamma \hat{\bm{\vartheta}}^{[m-1]} | \D)$ &  & \\
  $\hat{\bm{\vartheta}}^{[m]} = \gamma \hat{\bm{\vartheta}}^{[m-1]} + \nu \bm{u}^{[m]}$ & $\Rightarrow$ & ??? \\
  $\tbh^{[m+1]} = \tbh^{[m]} + \hat{\bm{\vartheta}}^{[m]}$ & &
\end{tabular}}
\vspace{0.2cm}
```

$\Rightarrow$ __Idea:__ Use Nestervos momentum and adjust it for functional updates and CWB.

## Nesterovs momentum in function space

- \citet{biau2019accelerated} proposed an adjustment of gradient boosting with Nesterovs momentum as optimizer.
- \citet{lu2020accelerating} lined out that this approach may diverge and proposed an corrected algorithm Accelerated Gradient Boosting Machine (AGBM):
```{=latex}
\vspace{-0.9cm}
\begin{align*}
g^{[m]} &= (1 - \theta_m) f^{[m]} + \theta_m h^{[m]}\\
f^{[m+1]} &= g^{[m]} + \eta b^{[m]} \\
h^{[m+1]} &= h^{[m]} + \eta / \theta_m b^{[m]}_{\text{cor}}
\end{align*}
```

with $\theta_m = 2 / (m + 1)$ and $h^{[m]}$ the momentum sequence.

## Base learners in AGBM

- $b^{[m]}$ is fit to pseudo residuals $\rmm$ w.r.t. $\hat{g}^{[m-1]}$ instead of $\fmh$
- $b^{[m]}_{\text{cor}}$ is fit to error-corrected pseudo residuals: $$c^{[m](i)} = \rmi + \frac{m}{m+1}(c^{[m-1](i)} - \hat{b}_{\text{cor}}^{[m-1]}(\xi)),$$ with $i = 1, \dots, n$, if $m > 1$ and $\bm{c}^{[m]} = \rmm$ if $m = 0$.



## Adapting AGBM for CWB

In \citet{schalk2022accelerated}, we proposed an accelerated CWB (ACWB) version by incorporating these adaptions to CWB, therefore:

- Both base learners, $b^{[m]}$ and $b^{[m]}_{\text{cor}}$, are the result of a selection process that chooses one of $b_1, \dots, b_K$ w.r.t. to the minimal SSE on the respective pseudo residuals $\rmm$ and $\bm{c}^{[m]}$.
- Update the estimated parameters accordingly to allow the estimation of partial feature effects.

Considering these issues allows to maintain all advantages of CWB in ACWB. We refer to the publication for details about the algorithms.

## Hybrid CWB

- We observed, that ACWB can overfit if not stopped early.
- Therefore, we combined ACWB with CWB to accelerate the fitting in the beginning and fine-tune the model using CWB:

\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{figures/fig-optim-emp-risk.pdf}
\end{figure}

# Memory

## Base learner design matrix

- Each base learner $b_1, \dots, b_K$ requires to build a design matrix $\bm{Z}_k\in\mathbb{R}^{n\times d_k}$ based on the feature vector $\bm{x}_k$.
- For example:

\begin{minipage}{0.75\textwidth}
{\tiny $
\underbrace{\begin{pmatrix} 80 \\ 31 \\ \vdots \\ 28 \\ 49 \end{pmatrix}}_{= \bm{x}_{\text{age}}}
$}
{\Large $\Rightarrow$}
{\tiny $
\underbrace{\begin{pmatrix}
    0 & 0 & 0 & 0.04 & 0.56 & 0.39 & 0.01  \\
    0.02 & 0.45 & 0.51 & 0.03 & 0 & 0 & 0 \\
    & & & \vdots & & & \\
    0.13 & 0.66 & 0.21 & 0 & 0 & 0 & 0 \\
    0 & 0 & 0.03 & 0.53 & 0.43 & 0.01 & 0
\end{pmatrix}}_{=\bm{Z}_{\text{age}}\ \text{(B-spline basis)}}
$}\hspace{0.2cm}{\Large $\Rightarrow$}
\end{minipage}
\begin{minipage}{0.2\textwidth}
\phantom{a}\vspace{-0.2cm}\hspace*{0.5cm}
```{r, echo=FALSE, fig.width=2, fig.height=1.6, out.width="3cm"}
ggage
```
\end{minipage}
\begin{itemize}
\item[$\Rightarrow$] If $n$ is large, the memory gets filled very fast (especially if $p$ is also large).
\end{itemize}

## Binning

- To reduce the memory consumption, we applied binning to operate on a reduced representation of $\bm{Z}_k$.
- Binning is a technique that allows to represent the $n$ values $x_k^{(1)}, \dots, x_k^{(n)}$ of $\xv_k$ by $n^\ast < n$ design points $\bm{z}_k = (z_k^{(1)}, \dots, z_k^{(n^\ast)})$.
- The idea is to assign each $x_k^{(i)}$ to the closest design point $z_k^{(i)}$ and store the assignment in a map $\text{ind}_k^{(i)}$: $x_k^{(i)} \approx z_k^{(\text{ind}_k^{(i)})}$

```{r, echo=FALSE, fig.width=4, fig.height=0.8, out.width="9cm", fig.align="center"}
xager = sample(xage, 20)
zage = seq(min(xager), max(xager), length.out = 5)
xagerb = vapply(xager, FUN.VALUE = numeric(1), FUN = function(x) zage[which.min(abs(x - zage))])

dfpl = data.frame(x = xager, z = xagerb)
dfd = data.frame(x = zage)

ggplot() +
  geom_segment(data = dfpl, aes(x = x, xend = z, y = 1, yend = 0), alpha = 0.5) +
  geom_point(data = dfpl, aes(x = x, y = 1)) +
  geom_point(data = dfpl, aes(x = z, y = 0)) +
  geom_point(data = dfd, aes(x = x, y = 0)) +
  theme_minimal() +
  xlab("Age") + ylab("") +
  scale_y_continuous(limits = c(-0.2, 1.2),
    breaks = c(0, 1),
    labels = expression(paste("Design points ", z[k]), paste("Feature values ", x[k])))

```


## Binning in GLMs

- \citet{lang2014multilevel} used binning to discretize feature vectors to increase the efficiency of multilevel structured additive regression.

- \citet{wood2017gigadata} applied binning to fit GAMs to gigadata and argue that the best approximation is achieved by setting $n^\ast = \sqrt{n}$.
- \citet{li2020faster} presented optimized cross-product operations of binned design matrices to also speed up the fitting.

## Binning in CWB

- Represent numerical features $\bm{x}_k$ by $n^\ast$ design points $\bm{z}_k$.
- Build the design matrix $\bm{Z}_k^\ast$ based on $\bm{z}_k$ which requires to store $n^\ast d_k$ values instead of $nd_k$.
- Use optimized cross-product operations to estimate the parameters $\tbh_k^{[m]}$ of base learner $b_k$ to also speed up the fitting.

\begin{minipage}{0.75\textwidth}
{\tiny $
\underbrace{\begin{pmatrix} 15 \\ 32 \\ 50 \\ 67 \\ 84 \end{pmatrix}}_{= \bm{z}_{\text{age}}}
$}
{\Large $\Rightarrow$}
{\tiny $
\underbrace{\begin{pmatrix}
    0.17 & 0.67 & 0.17 & 0    & 0    & 0    & 0    \\
    0    & 0.34 & 0.59 & 0.06 & 0    & 0    & 0    \\
    0    & 0    & 0.01 & 0.43 & 0.53 & 0.01 & 0    \\
    0    & 0    & 0    & 0.06 & 0.59 & 0.34 & 0    \\
    0    & 0    & 0    & 0    & 0.17 & 0.67 & 0.17
\end{pmatrix}}_{=\bm{Z}^\ast_{\text{age}}\ \text{(B-spline basis)}}
$}\hspace{0.2cm}{\Large $\Rightarrow$}
\end{minipage}
\begin{minipage}{0.2\textwidth}
\phantom{a}\vspace{-0.2cm}\hspace*{0.5cm}
```{r, echo=FALSE, fig.width=2, fig.height=1.6, out.width="3cm"}
ggage
```
\end{minipage}


# Application to data

## Benchmark

\begin{figure}
\centering
\includegraphics[width=1.1\textwidth]{figures/fig-eq2-2.pdf}
\end{figure}

## Binning with big data

- HIGGS: $11 \times 10^6$ observations, $29$ features, 2.4 GB
- NYC Yellow Taxi Trip: $24.3 \times 10^6$ observations, $22$ features, 3.3 GB
- FEMA’s National Flood Insurance Policy Database: $14.5 \times 10^6$ observations, $50$ features, 3.4 GB

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/app-big-data.pdf}
\end{figure}

## {.plain}

## {.allowframebreaks .plain}

\footnotesize
\bibliographystyle{apalike}
\bibliography{references.bib}

